## Identification by Fixing Legislator's Signs

We can identify the scale and location of the latent dimensions by fixing the mean and location of the distribution of the legislator's ideal points.
This can be done by:

$$
\xi_i \sim \mathsf{Normal}(0, 1)
$$

This does not exactly fix the mean of $\xi$ in any particular simulation. As the sample size increases, $n \to \infty$ will give $mean(\xi) \to 0$ and $sd(\xi) \to 1$. This "soft-identification" will also overestimate the uncertainty in the ideal points of legislators (see paper by ... ? ). So in the estimation, I do the following to ensure that in each simulation, $\xi$ has exactly mean 0 and standard deviation 1.

$$
\begin{aligned}[t]
\xi_i^{*} &\sim \mathsf{Normal}(0, 1) \\
\xi_i &= \frac{\xi^*_i - \mathrm{mean}(\xi)}{\mathrm{sd}(\xi)}
\end{aligned}
$$

However, this does not identify the direction of the latent variables. This can be done by restriction the sign of either a single legislator ($\xi_i$) or roll-call ($\beta_i$).

In this case, instead of fixing the ideal points of two legislators, instead use skew-normal distributions to set the sides of each legislator.

$$
\beta_j \sim \mathsf{SkewNormal}(0, 2.5, d_j)
$$

where

$$
d_j =
\begin{cases}
-50 & \text{if Democratic party line vote} \\
0 & \text{not a party line vote} \\
50 & \text{if Republican party line vote}
\end{cases}
$$

The skew-normal distribution is an extension of the normal distribution with an additional skewness parameter,

$$
y \sim \mathsf{SkewNormal}(\mu, \sigma, \alpha)
$$

as $|\alpha| \to \infty$, the skew-normal approaches a half-normal distribution.

```{r}
map_df(
  c(-50, 0, 50),
  function(alpha) {
    tibble(
      x = seq(-4, 4, by = 0.1),
      density = dsn(x, alpha = alpha),
      alpha = alpha
    )
  }
) |>
  ggplot(
    aes(
      x = x,
      y = density,
      colour = factor(alpha)
    )
  ) +
  geom_line()
```

```{r}
mod_ideal_point_3 <- stan_model("stan/ideal_point_3.stan")
```


```{r}
mod_ideal_point_3
```

``` {.default}
// ideal point model
// identification:
// - ideal points ~ normal(0, 1)
// - signs of ideal points using skew normal
data {
  // number of individuals
  int N;
  // number of items
  int K;
  // observed votes
  int<lower = 0, upper = N * K> Y_obs;
  int y_idx_leg[Y_obs];
  int y_idx_vote[Y_obs];
  int y[Y_obs];
  // priors
  // on items
  real alpha_loc;
  real<lower = 0.> alpha_scale;
  real beta_loc;
  real<lower = 0.> beta_scale;
  // on ideal points
  vector[N] xi_skew;
}
parameters {
  // item difficulties
  vector[K] alpha;
  // item discrimination
  vector[K] beta;
  // unknown ideal points
  vector[N] xi_raw;
}
transformed parameters {
  // create xi from observed and parameter ideal points
  vector[Y_obs] mu;
  vector[N] xi;

  xi = (xi_raw - mean(xi_raw)) ./ sd(xi_raw);
  for (i in 1:Y_obs) {
    mu[i] = alpha[y_idx_vote[i]] + beta[y_idx_vote[i]] * xi[y_idx_leg[i]];
  }
}
model {
  alpha ~ normal(alpha_loc, alpha_scale);
  beta ~ normal(beta_loc, beta_scale);
  // soft center ideal points
  // in transformed block enforce hard-centering
  xi_raw ~ skew_normal(0., 1., xi_skew);
  y ~ bernoulli_logit(mu);
}
generated quantities {
  vector[Y_obs] log_lik;

  for (i in 1:Y_obs) {
    log_lik[i] = bernoulli_logit_lpmf(y[i] | mu[i]);
  }
}
```

Instead of fixing the ideal points, set the skewness parameter of the skew normal distribution so that $\xi_{\text{FRIST (R TN)}} > 0$:

```{r}
legislators_data_2 <- within(list(), {
    y <- as.integer(s109_votes$yea)
    y_idx_leg <- as.integer(s109_votes$.legis_id)
    y_idx_vote <- as.integer(s109_votes$.rollcall_id)
    Y_obs <- length(y)
    N <- max(s109_votes$.legis_id)
    K <- max(s109_votes$.rollcall_id)
    # priors
    alpha_loc <- 0
    alpha_scale <- 5
    beta_loc <- 0
    beta_scale <- 2.5
    xi_skew <- if_else(s109_legis_data$legislator == "FRIST (R TN)", 50, 0)
  }
)
```

```{r}
legislators_init_2 <- function(chain_id) {
  list(
    xi_raw = if_else(
      s109_legis_data$party == "Republican",
      1,
      if_else(
        s109_legis_data$party == "Democratic",
        -1,
        0
      )
    )
  )
}
```


```{r}
legislators_fit_2 <- sampling(
  mod_ideal_point_3,
  data = legislators_data_2,
  init = legislators_init_2,
  chains = 1, iter = 500,
  pars = c("alpha", "beta", "xi")
)
```

## Identification by Discrimination Parameters' Signs

Alternatively, we can identify the location and scale of the latent dimensions with the legislator's ideal points,

$$
\begin{aligned}[t]
\xi_i^{*} &\sim \mathsf{Normal}(0, 1) \\
\xi_i &= \frac{\xi^{*}_i - \mathrm{mean}(\xi)}{\mathrm{sd}(\xi)} ,
\end{aligned}
$$

and identify the rotation of latent dimensions by fixing the sign of the discrimination parameter of a single roll-call vote ($\beta_j$).

```{r}
# mod_ideal_point_2 <- stan_model("stan/ideal_point_2.stan")
```


```{r}
# mod_ideal_point_2
```

```{.default}
// ideal point model
//
// identification:
// - ideal points ~ normal(0, 1)
// - signs of item discrimination using skew normal
data {
  // number of individuals
  int N;
  // number of items
  int K;
  // observed votes
  int<lower = 0, upper = N * K> Y_obs;
  int y_idx_leg[Y_obs];
  int y_idx_vote[Y_obs];
  int y[Y_obs];
  // priors
  // on items
  real alpha_loc;
  real<lower = 0.> alpha_scale;
  vector[K] beta_loc;
  vector<lower = 0.>[K] beta_scale;
  vector[K] beta_skew;
}
parameters {
  // item difficulties
  vector[K] alpha;
  // item discrimination
  vector[K] beta;
  // unknown ideal points
  vector[N] xi_raw;
}
transformed parameters {
  // create xi from observed and parameter ideal points
  vector[Y_obs] mu;
  vector[N] xi;
  xi = (xi_raw - mean(xi_raw)) ./ sd(xi_raw);
  for (i in 1:Y_obs) {
    mu[i] = alpha[y_idx_vote[i]] + beta[y_idx_vote[i]] * xi[y_idx_leg[i]];
  }
}
model {
  alpha ~ normal(alpha_loc, alpha_scale);
  beta ~ skew_normal(beta_loc, beta_scale, beta_skew);
  // soft center ideal points
  // in transformed block enforce hard-centering
  xi_raw ~ normal(0., 1.);
  y ~ bernoulli_logit(mu);
}
generated quantities {
  vector[Y_obs] log_lik;
  for (i in 1:Y_obs) {
    log_lik[i] = bernoulli_logit_lpmf(y[i] | mu[i]);
  }
}
```

As before, we will restrict the sign using a skew-normal distribution with a large skewness parameter.

Theoretically, it does not matter which parameter is restricted, but it can be useful for both interpretation and computation to restrict the sign of the parameter that *ex ante* you expect to be far from zero.

Since, in this case, it is expected that the primary dimension is liberal-conservative, and that the current Republican/Democratic parties divide on those lines, we will choose a bill that splits on party lines.

We'll fix $\beta_{\text{2-169}}$ which was a roll-call vote which perfectly split on party lines: 55 Republican senators in favor, and 43 Democratic senators opposed. Of the roll-call votes that perfectly split on party lines, this had the most total votes cast.

```{r}
legislators_data_3 <- within(list(), {
    y <- as.integer(s109_votes$yea)
    y_idx_leg <- as.integer(s109_votes$.legis_id)
    y_idx_vote <- as.integer(s109_votes$.rollcall_id)
    Y_obs <- length(y)
    N <- max(s109_votes$.legis_id)
    K <- max(s109_votes$.rollcall_id)
    # priors
    alpha_loc <- 0
    alpha_scale <- 5
    beta_loc <- rep(0, K)
    beta_scale <- rep(2.5, K)
    beta_skew <- if_else(s109_vote_data$rollcall == "2-169", 50, 0)
  }
)
```

```{r}
legislators_init_3 <- function(chain_id) {
  list(
    beta = if_else(
      s109_vote_data$partyline %in% "Republican",
      1,
      if_else(
        s109_vote_data$partyline %in% "Democratic",
        -1,
        0
      )
    ),
    alpha = plogis(s109_vote_data$yea_pct),
    xi_raw = if_else(
      s109_legis_data$party == "Republican",
      1,
      if_else(
        s109_legis_data$party == "Democratic",
        -1,
        0
      )
    )
  )
}
```

```{r}
# legislators_fit_3 <- sampling(
#   mod_ideal_point_2,
#   data = legislators_data_3,
#   init = legislators_init_3,
#   chains = 1, iter = 500,
#   pars = c("alpha", "beta", "xi")
# )
```

## Questions

-   Plot the the distributions of $\alpha$ and $\beta$.

-   Estimate the model with improper priors for $\alpha$, $\beta$, and $\xi$. What happens?

-   Estimate the the model that fixes the distribution of legislators, but
    does not fix the signs of legislators. Run two chains. In one chain use
    the following starting values, $\xi_i = 1$ for all Democratic and
    Independent Senators, $\xi_i = -1$ for all Republican Senators; in the
    other use $\xi_i = -1$ for all Democratic and Independent Senators, and
    $\xi_i = 1$ for all Republican Senators. Visualize the densities of
    various $\xi_i$ points. Look for evidence of bimodality.

-   Extend the model to $K > 1$ dimensions.

[^legislators-src]: Example derived from Simon Jackman, "Legislators: estimating legislators' ideal points from voting histories (roll call data)", *BUGS Examples,* 2007-07-24, [URL](https://web-beta.archive.org/web/20070724034141/http://jackman.stanford.edu:80/mcmc/legislators.odc).
