# %load https://raw.githubusercontent.com/jpowerj/dsan-content/refs/heads/main/2025-sum-dsan5650/submit_demo.py
import datetime, getpass, ipylab, ipywidgets, IPython.display, pytz, time
jfe = ipylab.JupyterFrontEnd()
button, output = ipywidgets.Button(description="Submit HW1"), ipywidgets.Output()
IPython.display.display(button, output)
def generate_timestamp(datetime_obj):
    return str(datetime_obj).split(".")[0].replace(" ","_").replace(":","").replace("-","")
def on_button_clicked(b):
    with output: print("Saving notebook...")
    jfe.commands.execute('docmanager:save')
    time.sleep(0.2)
    output.clear_output()
    with output: print("Submitting...")
    cur_time = datetime.datetime.now(pytz.timezone('US/Eastern'))
    cur_ts = cur_time.strftime('%X %p, %x')
    cur_ts_short = generate_timestamp(cur_time)
    !/srv/submit.sh HW1 {str(cur_ts)}
    time.sleep(0.2)
    output.clear_output()
    with output: print(f"HW1 submitted at {str(cur_ts)} [Check feedback folder]")
    netid = getpass.getuser().replace("jupyter-","")
    time.sleep(1)
    jfe.commands.execute('filebrowser:open-path', args={'path': f'HW1/feedback/{cur_ts_short}_{netid}.txt'})
button.on_click(on_button_clicked)

---

## Proxies for $Z$

```{r, engine = 'tikz'}
\usetikzlibrary{calc,intersections,through,backgrounds,bayesnet}
\begin{tikzpicture}[scale=1.0]
\node[draw, circle] (X) {$X$};
\node[draw, circle, right=of X] (Y) {$Y$};
\path[->] (X) (Y);
\end{tikzpicture}
```

---

```{mermaid}
%%| width: 340
%%| fig-align: center
flowchart TB
  P{PGMs}
  P --> D[Use for Computational<br>Social Science]
  P --> E[Extend to Causal Inference]
```

---

## Our First DAGventure (sry)

```{r}
#| label: first-dag
library(rethinking)
library(dagitty)
dag5.1 <- dagitty("dag{ A -> D; A -> M; M -> D;}")
coordinates(dag5.1) <- list(x=c(A=0,D=1,M=2), y=c(A=0, D=1, M=0))
drawdag(dag5.1)
```
